---
title: "ddRAD_analysis"
output: github_document
---

#This script reads in data files from the ipyrad output and subsets the data to get only ddRAD data and extract only SNPs without missing data.

Set directory and read in .str datafile from iPyrad output. This is the unlinked SNPs file.

```{r}
data2.2 <- read.delim("data3.u.str", header=FALSE)
data <- data2.2[,colSums(is.na(data2.2))<nrow(data2.2)]
```

Vector of  unlinked SNP IDs corresponding to data3.snps.map

```{r}
unlinkedsnps <- seq(c(1:11822))
data2 <- data[,2:11823]
colnames(data2) <- unlinkedsnps
data3 <- cbind(data[,1],data2)
data4 <- t(data3)
data5 <- as.data.frame(data4)
```

Extract only ddRAD data 

```{r}
data6 <- data5[,c(1,2,5,6,9,10,13,14,17,18,21,22,25,26,29,30,33,34,37,38,41,42,47,48,51,52,55,56,59,60,63,64,69,70,73,74,77,78,81,82,85,86,89,90,93,94,97,98,101,102,105,106,109,110)]
```

Remove any SNPs with missing data (-9 is the NA value)

```{r}
data7 <- data6[!rowSums(data6 == -9) >= 1,]
data8 <-t(data7)
rownames(data8) <- data8[,1]
data9 <- data8[,2:1114]
write.table(data9, file = "data3-2.str", row.names = TRUE, col.names = TRUE, quote = FALSE)
```

Read in a .geno file from the ipyrad output and extract ddRAD data without missing values. This is the best filetype for MDS.

```{r}
a1 <- read.table("data3.u.geno", colClasses = 'character', header = FALSE)
a2 <- read.fwf("data3.u.geno", widths=rep(1, max(nchar(a1$V1))), colClasses = 'numeric', header=FALSE)
header <- read.delim("header_data3.txt", header=FALSE)
names <- t(header)
names2 <-as.vector(names)
colnames(a2) <- names2
```

Select samples of interest (some have very low sample sizes)

```{r}
a3 <- a2[,c(1,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,24,25,26,27,28,29,30,31,32,33,35,36,37,38,39,40,41,42,43,44,45,46,47,48,49,50,51,52,53,54,55,56)]
```

Matrix with only ddr loci **if including sample 101

```{r}
a4 <- a3[,c(1,2,4,6,8,10,12,14,16,18,20,22,24,26,28,30,32,34,36,38,40,42,44,46,48,50,52)]
```

Get rid of rows with any NAs (9)

```{r}
a5 <- a4[!rowSums(a4 == 9) >= 1,]
head(a5)
```

Read in sample info and transpose. Columns are sample, depth, type, symbiont type, and branch diamater.

```{r}
sinfo <- read.table("sample_info.txt", colClasses = 'character', header = FALSE)
head(sinfo)
tsinfo <- t(sinfo)
```

Create vectors for morphotype and diameter (note whether sample 101 was included or not)

```{r}
type <- tsinfo[3,]
diam <- tsinfo[5,]
```

MDS of SNPs

```{r}
a6 <- t(a5)
d <- dist(a6) # euclidean distances between the rows
fit <- cmdscale(d,eig=TRUE, k=2)
x <- fit$points[,1]
y <- fit$points[,2]
```

This adds a column of color values based on branch diameter

```{r}
data_seq = seq(min(as.numeric(diam)), max(as.numeric(diam)), length=25)
col_pal = colorRampPalette(c('blue', 'green', 'red'))(25+1)
cols = col_pal[ cut(as.numeric(diam), data_seq, include.lowest=T) ]
```

Plot MDS and legend

```{r}
layout(matrix(1:2,ncol=2), width = c(3,1),height = c(1,1))
plot(x, y, xlab="Coordinate 1", ylab="Coordinate 2", 
     main="MDS colored by branch diameter",  type="n")
points(x,y, col = cols, pch = 19)

legend_image <- as.raster(matrix(sort(cols, decreasing = TRUE), ncol=1))
plot(c(0,2),c(0,1),type = 'n', axes = F,xlab = '', ylab = '', main = 'branch diamater')
text(x=1.5, y = seq(0,1,l=5), labels = seq(6,26,l=5))
rasterImage(legend_image, 0, 0, 1,1)
```

DAPC using adegenet

```{r message=FALSE}
library("adegenet")
library("ade4")
```

Read in unlinked SNP file created at the top of this script.

```{r}
genind1 <- read.structure("data3-2.str", n.ind = 27, n.loc = 1113, 
                          onerowperind = FALSE, col.lab = 1, 
                          NA.char = "-9", ask = FALSE, 
                          row.marknames = 1, quiet = FALSE)
```

Find optimal number of clusters irrespective of species id. In this case, best to retain all PCs. 

```{r}
groups <- find.clusters(genind1, max.n.clust=10, n.pca = 24,
                        choose.n.clust = FALSE, criterion = "min")
```

Cross validation to determine number of PCs to retain.

```{r}
xval <- xvalDapc(genind1@tab, groups$grp, n.pca.max = 25, training.set  = 0.9,result = "groupMean", center = TRUE, scale = FALSE,n.pca = NULL, n.rep = 100, xval.plot = TRUE)
```

Show max number of PCs to retain

```{r}
xval[2:6]
```

Perform dapc using groups defined above group (groups$grp). Note that n.pca and n.da can be left blank and the program will query which values to choose.


```{r}
dapc1 <- dapc(genind1, pop = groups$grp, n.pca=9, n.da=2)
```

Scatterplot of results

```{r}
scatter(dapc1, #label.inds = list(air = 0.05, pch = 0.5),
        posi.da = "topleft", posi.pca = "topright")
```

Look at loadings of individual loci

```{r}
set.seed(4)
contrib <- loadingplot(dapc1$var.contr, axis=1, 
                       thres=.005, lab.jitter=1)
```

Boxplot comparing branch diameter among groups from dapc

```{r}
diam2 <- as.numeric(sinfo[,5])
boxplot(diam2 ~ dapc1$assign, xlab = "Group", ylab =  "diameter")
```

ANOVA of diameter by group

```{r}
#Run lm on diameter by group
model <- lm(diam2 ~ dapc1$assign)
#Check model (qq plot, etc)
par(mfrow=c(2,2))
plot(model)
library(lmtest)
#Breush Pagan Test for heteroscadisticity
bpt <- bptest(model)
print(bpt)
#Run ANOVA
aov <- anova(model)
print(summary(aov))
#pairwise t test with bonferonni adjustment
ttest <- pairwise.t.test(diam2, dapc1$assign, p.adj = "bonf")
print(ttest)
```

Multiple regression: test linear model of depth, symbiont type, and
branch diameter vs. dapc of SNPs using a single DA from the model above

```{r}
#dapc with single DF
dapc1 <- dapc(genind1, pop = groups$grp, n.pca=9, n.da=1)
#Select # pcs = 9, # df = 1
scatter(dapc1)
#convert DF coord to vector
dapc1_da1 <- dapc1$ind.coord
#replace unknown symbiont type NAs to "U"
sinfo[is.na(sinfo)] <- "U"
#fit model with water depth (sinfo$V2), branch diameter (sinfo$V5), 
# and symbiont type (sinfo$V4) 
fit <- lm(na.omit(dapc1_da1 ~ as.numeric(sinfo$V2) + as.factor(sinfo$V4) + as.numeric(sinfo$V5)))
#Relative importance of different variables in model
library(relaimpo)
relimp <- calc.relimp(fit,type=c("lmg","last","first"),
            rela=TRUE)
print(relimp)
```